Assessing linkages between watershed nutrient loading and water quality in a subtropical estuary with semiparametric models.

### 2023-05-20 Schramm-2023-05-AS/Schramm-2023-05-AS.Rmd
init springer nature tex submission for AS

---
title: "Assessing linkages between watershed nutrient loading and water quality in a subtropical estuary with semiparametric models."
format:
  wordcount-docx:
    toc: false
    number-sections: false
    reference-doc: format.docx
author: 
  - name: Michael Schramm
    email: michael.schramm@ag.tamu.edu
    corresponding: true
    affiliations:
      - ref: agrilife
affiliations:
  - id: agrilife
    name: Texas A&M AgriLife Research, Texas Water Resources Institute
    city: College Station
    state: Texas
    country: United States of America
crossref:
  fig-prefix: Fig   # (default is "Figure")
  tbl-prefix: Table    # (default is "Table")
  title-delim: .     # (default is ":")
  ref-hyperlink: false # (default is true)
  eq-prefix: Eq
bibliography: mybib.yaml
csl: https://www.zotero.org/styles/springer-basic-author-date-no-et-al?source=1
filters: 
  - authors-affiliations.lua
keep-md: true
cite-method: citeproc
link-bibliography: false
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(targets)
library(knitr)
library(tidyverse)
library(kableExtra)
library(patchwork)
library(ggrepel)
library(rcartocolor)
library(twriTemplates)
library(ggspatial)
library(ragg)
library(flextable)
library(modelsummary)
library(scico)
library(ggridges)
library(ggtext)
library(lubridate)
library(units)
library(gratia)
library(gt)
library(officer)

ragg_tiff <- function(..., res = 300) {
  ragg::agg_tiff(..., res = res, units = "in")
}

## this sets our default code chunk options
knitr::opts_chunk$set(echo = FALSE,
                      ## figure chunk options
                      dev = "ragg_tiff",
                      dpi = 300,
                      fig.align = "center",
                      fig.ext = "tiff",
                      out.width = "100%")

@@

# Abstract
Lavaca Bay is a small secondary embayment on the Texas coast that is
displaying early signals of water quality degradation. This
study applied a semiparametric approach to assess both watershed nutrient loads 
and the responses in estuary water quality to nutrient loading and streamflow. 
Cross-validation indicated that, despite data constraints, semiparametric models 
performed well at nutrient load prediction. Based on these models, delivered
annual nutrient loads varied substantially from year to year. In contrast,
minimal changes in calculated flow-normalized loads indicate that changes in 
nutrient loading were driven by natural variation in precipitation and runoff as
opposed to changes in nonpoint sources.
Estuary water quality models did not identify significant long-term changes 
within Lavaca Bay for dissolved oxygen or chlorophyll-*a*. However, site specific 
long-term increases in both organic and inorganic nitrogen are concerning. 
Further analysis found freshwater inflow was a strong driver of nutrient and 
chlorophyll-*a* concentrations within Lavaca Bay but there was no evidence that 
changes in nutrient loading explained variation in dissolved oxygen or 
chlorophyll-*a* concentrations. In addition to providing a baseline assessment 
of watershed nutrient loading and water quality responses in the Lavaca Bay 
watershed, this study provides methodological support for the use of 
semiparametric methods in load regression models and estuary assessments.

# Introduction

@@

# Materials and Methods

## Study Area and Data

Lavaca Bay is 190 km^2^ with the majority of freshwater inflow provided by the 
Lavaca and Navidad River systems (@fig-fig1). The Garcitas-Arenosa, 
Placedo Creek, and Cox Bay watersheds provide additional freshwater inflows. 
The entire watershed land area is 8,149 km^2^ and primarily rural. Watershed 
land cover and land use is 50% grazed pasture and rangeland, 20% cultivated 
cropland (primarily rows crops such as corn, cotton, and sorghum), and 5% 
suburban/urban. Pasture and rangeland is concentrated in the Lavaca River 
watershed, while cultivated crops are generally located along the eastern 
tributaries of the Navidad river. The Lavaca and Navidad River watersheds are a 
combined 5,966 km^2^, or approximately 73% of the entire Lavaca Bay watershed 
area. Discharge from the Navidad River is regulated by Lake Texana which has 
been in operation since 1980. Lake Texana provides 0.210 km^3^ of water 
storage and discharges into the tidal section of the Navidad River which 
ultimately joins the tidal section of the Lavaca River 15 km upstream of the 
confluence with the Lavaca Bay.

```{r}
#| label: fig-fig1
#| fig-width: 5.2
#| fig-height: 6.75
#| out-width: "5.2in"
#| fig-cap: "Map of Lavaca Bay and the contribution watershed. The freshwater sites are the most downstream freshwater stream locations with water quality and streamflow data used for nutrient load models. Water quality concentration data at the three Lavaca Bay sites were used to assess relationships between freshwater flows, loads and estuary water quality."

knitr::include_graphics("Figures/fig1-1.tiff")

@@

Daily discharges for the Lavaca River (USGS-08164000, @fig-fig1) 
were obtained from the United States Geologic Survey (USGS) National Water 
Information System using the *dataRetrieval* R package 
[@deciccoDataRetrievalPackagesDiscovering2022]. Gaged daily discharges from 
the outlet of Lake Texana on the Navidad River (USGS-0816425) were provided by 
the Texas Water Development Board (TWDB) 
(April 21, 2022 email from R. Neupane, TWDB).

Water quality sample data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. 
The QAPP and procedures manuals ensure the consistent collection and laboratory 
methods are applied between samples collected by different entities and under 
different projects. 
All sites had varying lengths of and availability of data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (`#r run_reference("table1")`). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper Lavaca Bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site (TCEQ-13383), and the 
lower Lavaca Bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO*~x~*), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(@tbl-table2).

```{r}

@@

datasummary(`Station ID` * (`TP (mg/L)` + `NO3 (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
               Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
            data = df, output = "kableExtra", format = "simple")

# datasummary(`Station ID` * (`TP (mg/L)` + `NO3 (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
#                Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
#             data = df, output = "flextable") |> 
#   set_caption(caption = "Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.",
#               autonum = run_autonum(seq_id = "tab", bkm = "table1")) |> 
#   style(pr_t = fp_text_default(font.size = 9), part = "body") |>
#   style(pr_t = fp_text_default(font.size = 9, bold = TRUE), part = "header") |>
#   set_table_properties(layout = "autofit") |>
#   padding(padding.top = 0, padding.bottom = 0, part = "body")
```

```{r}
#| warning: false

@@

modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                                            `NOx (mg/L)` +
                                            `TKN (mg/L)` +
                                            `Chlorophyll-a (ug/L)` +
                                            `DO (mg/L)`) ~
                            Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                          data = df, output = "flextable") |> 
  set_caption(caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.",
              autonum = run_autonum(seq_id = "tab", bkm = "table2")) |> 
  style(pr_t = fp_text_default(font.size = 9), part = "body") |>
  style(pr_t = fp_text_default(font.size = 9, bold = TRUE), part = "header") |>
  set_table_properties(layout = "autofit") |>
  padding(padding.top = 0, padding.bottom = 0, part = "body")

 # modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
 #                            `NOx (mg/L)` +
 #                            `TKN (mg/L)` +
 #                            `Chlorophyll-a (ug/L)` +
 #                            `DO (mg/L)`) ~
 #                            Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
 #                          data = df, output = "kableExtra", format = "pipe") #|>
  #  flextable::autofit()
   # mutate(`Station ID` = case_when(
   #   `Station ID` == "" ~ NA_character_,
   #   `Station ID` != "" ~ `Station ID`
   #   )) |>
   # tidyr::fill(`Station ID`, .direction = "down") |> 
   # gt::gt() |> 
   # gt::as_word()

@@

Estimates of NO~3~ and TP loads at the Lavaca River (USGS-08164000) and the 
outlet of Lake Texana on the Navidad River (USGS-08164525) were developed using 
GAMs relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to predict
nutrient concentrations for each day in the study period. GAMs were fit using 
the *mgcv* package in R which makes available multiple 
types of smooth functions with automatic smoothness selection [@woodFastStableRestricted2011]. 
The general form of the model related NO~3~ or TP concentration to a long term
tend, season, streamflow, and two different antecedent discharge terms, 
shown in [@eq-eq1]:

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(log1p(Q)) + f_4(ma) + f_5(fa)  \\
y &\sim \mathcal{N}(\mu,\,\sigma^{2})
\end{align*}
$${#eq-eq1}

where *μ* is the conditional expected NO~3~ or TP concentration, *g()* is the 
log-link, *α* is the intercept, *f~n~()* are smoothing functions. *y* is the 
response variable (NO~3~ or TP concentration) modeled as normally distributed 
with mean *μ* and standard deviation *σ*. *ddate* is the date converted to 
decimal notation, *yday* is numeric day of year (1-366), and *log1p(Q)* is the 
natural log of mean daily streamflow plus 1.

Moving average (*ma*) is an exponentially smoothed moving average that attempts 
to incorporate the influence of prior streamflow events on concentration at the 
current time period [@wangLoadEstimationUncertainties2011;  
@kuhnertQuantifyingTotalSuspended2012; @zhang_improving_2017], using [@eq-eq2]:

$$
\begin{align*}
ma(\delta) &= d{\kappa_{i-1}}+(1-\delta)\hat{q}_{i-1},\\
\kappa_{i} &=\sum_{m=1}^{i}\hat{Q}_m,
\end{align*}
$${#eq-eq2}

where *δ* is the discount factor (here, set equal to 0.95), κ~i~
is the cumulative flow (*Q*) up to the *i*th day.

Flow anomaly (*fa*) is a unitless term that represents how wet or dry the 
current time period is from a previous time period [@vecchiaTrendsConcentrationsUse2009; 
@zhang_improving_2017]. Long-term flow anomaly (*ltfa*) is the streamflow over 
the previous year relative to the entire period [@zhang_improving_2017] and 
calculated using [@eq-eq3]:

$$
ltfa(t) = \bar{x}_{1\,year}(t) - \bar{x}_{entire\,period}
$${#eq-eq3}

and the short-term flow anomaly (*stfa*) calculated as the current day flow 
compared to the preceding 1-month streamflow using [@eq-eq4]:

$$
stfa(t) = x_{current\,day}(t) - \bar{x}_{1\,month}(t)
$${#eq-eq4}

where *x* are the averages of log-transformed streamflow over the antecedent 
period (*1-year*, *1-month*, etc.) for time *t*.  We used *ltfa* in NO~3~ models 
and *stfa* in TP models based on previous work demonstrating major improvements 
in NO~x~ regression models that incorporated 
*ltfa* and moderate improvements in TP regression models that incorporated 
*stfa* [@zhang_improving_2017].

@@

To test if changes in freshwater inflow and nutrient loading had explanatory
effect on changes in estuary water quality a series of GAM models were fit at
each site relating parameter concentration to temporal trends ([@eq-eq5]), 
temporal trends and inflow (@eq-eq6), and temporal trends, inflow, and
nutrient loads (@eq-eq7):

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday)\\
y &\sim \Gamma(\mu,\lambda)
\end{align*}
$${#eq-eq5}

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(Q) \\
y &\sim \Gamma(\mu,\lambda)
\end{align*}
$${#eq-eq6}

$$
\begin{align*}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(Q) + f_4(Load) \\
y &\sim \Gamma(\mu,\lambda)
\end{align*}
$${#eq-eq7}

where *μ* is the conditional expected response (nutrient concentration), *g()* 
is the log link, and response variable was modeled as Gamma distributed with 
mean *μ* and scale *λ*. *f~1~(ddate)* is decimal date smoothed with a 
thin-plate regression spline, *f~2~(yday)* is the numeric day of year smoothed 
with a cyclic cubic regression spline, *f~3~(Q)* is mean daily inflow (the 
combined measurements from Lavaca River and Navidad River) and *f~4~(Load)* is 
the total NO~3~ or TP watershed load. The set of models specified for each water
quality response are in @tbl-table3.

```{r}
#| warning: false
df <- tibble(
  `Water Quality Response Parameter` = c(rep("TP", 3), rep("NOx", 3),
               rep("Chlorophyll-a", 3), rep("Dissolved Oxygen", 3),
               rep("TKN", 2)),
  Model = c("Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow"),
  `Model Terms` = c("s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(NO3 Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO3 Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO3  Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)")
)

# kbl(df, format = "pipe")

flextable::flextable(df) |> 
  set_caption(caption = "Set of GAM models specified for each water quality parameter response.",
              autonum = run_autonum(seq_id = "tab", bkm = "table3")) |> 
  style(pr_t = fp_text_default(font.size = 9), part = "body") |>
  style(pr_t = fp_text_default(font.size = 9, bold = TRUE), part = "header") |>
  set_table_properties(layout = "autofit") |>
  padding(padding.top = 0, padding.bottom = 0, part = "body")

@@

Because streamflow and nutrient loads are tightly correlated, freshwater inflow
can mask signals from nutrient loads alone. Following the 
methodology implemented by @murphyNutrientImprovementsChesapeake2022,
both streamflow and nutrient loads were prepossessed to account for season
and flow. Freshwater inflow and nutrient loads were replaced by 
seasonally adjusted log transformed inflow and flow-adjusted 
log transformed nutrient loads obtained by fitting a GAM relating season 
(day of year) to log transformed daily freshwater inflow values
and a GAM relating log transformed NO~3~ or TP loads to log transformed daily
inflow.Response residuals from the respective GAM models were 
used as *Q* and *Load* in @eq-eq6 and @eq-eq7.

@@

# Results

@@

Using evaluation criteria recommended by @moriasiHydrologicWaterQuality2015, 
predictive performance of nutrient loading GAMs ranged from "satisfactory" to
"very good" based on median NSE, *r*, and PBIAS metrics calculated using 5-fold 
cross validation. Median goodness-of-fit metrics for NO~3~ models in the Lavaca 
River were 0.34 NSE, 0.70 *r*, and 2.00 PBIAS. Navidad River NO~3~ models 
appeared to perform slightly better with 0.48 NSE and 0.87 *r* but with higher 
bias at 10.90 PBIAS. Generally, TP models performed better than NO~3~ models. 
Median goodness-of-fit metrics for TP in the Lavaca River were 0.81 NSE, 
0.93 *r*, and -7.20 PBIAS. Navidad River TP models has similar performance 
with 0.91 NSE, 0.99 *r*, and -3.30 PBIAS. Density plots of
metrics show similar distribution of values between sites for the 
same parameter, with the exception *r* values for NO~3~ loads where 
Lavaca River had a much larger variance in values compared to the Navidad River
(@fig-fig2). TP GAMS had higher average NSE and *r* values and 
lower variance in metric values compared to NO~3~.

```{r}
#| label: fig-fig2
#| warning: false
#| message: false
#| fig-width: 6.85
#| fig-asp: 0.65
#| out-width: "6.85in"
#| fig-cap: "Density plots of goodness-of-fit metrics (NSE, *r*, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calcualted from the empirical cumulative distribution of the goodness-of-fit metrics."

@@

Predicted annual NO~3~ and TP loads show considerable variation, generally following
patterns in discharge (@fig-fig3; @fig-fig4). 
Flow-normalized TP loads at both sites and flow-normalized NO~3~ loads in the Lavaca River
indicated watershed based loads did not change much over time
when accounting for variation driven by streamflow (@fig-fig3). 
Flow-normalized loads in the Lavaca River showed small variation over time with 
some decreases in NO~3~ loads since 2013.

>>>>

---
# Supported options: 
#   sn-nature:       Style for submissions to Nature Portfolio journals
#   sn-basic:        Basic Springer Nature Reference Style/Chemistry Reference Style
#   sn-mathphys:     Math and Physical Sciences Reference Style
#   sn-aps:          American Physical Society (APS) Reference Style
#   sn-vancouver:    Vancouver Reference Style
#   sn-apa:          APA Reference Style 
#   sn-chicago:      Chicago-based Humanities Reference Style
#   default:         Default

classoptions: 
  - sn-basic      
  # - Numbered      # Optional: Use numbered references instead of namedate references (only for sn-nature, sn-basic, sn-vancouver, sn-chicago, sn-mathphys or sn-nature)
  - referee       # Optional: Use double line spacing 
  - lineno        # Optional: Add line numbers
  # - iicol         # Optional: Double column layour

title: Assessing linkages between watershed nutrient loading and water quality in a subtropical estuary with semiparametric models.
titlerunning: Watershed nutrient loading and water quality in a subtropical estuary

authors: 
  - firstname: Michael
    lastname: Schramm
    email: michael.schramm@ag.tamu.edu
    affiliation: [1]
    corresponding: TRUE

affiliations:
  - number: 1
    corresponding: TRUE
    info:
      orgdiv: Texas Water Resources Institute
      orgname: Texas A&M AgriLife Research
    address:
        street: 1001 Holleman Dr. E.
        city: College Station
        postcode: 77840
        state: TX
        country: United States

keywords:
  - key
  - dictionary
  - word

# Sample for unstructured abstract 
abstract: |
  Lavaca Bay is a small secondary embayment on the Texas coast that is displaying early signals of water quality degradation. This study applied a semiparametric approach to assess both watershed nutrient loads and the responses in estuary water quality to nutrient loading and streamflow. Cross-validation indicated that, despite data constraints, semiparametric models performed well at nutrient load prediction. Based on these models, delivered annual nutrient loads varied substantially from year to year. In contrast, minimal changes in calculated flow-normalized loads indicate that changes in nutrient loading were driven by natural variation in precipitation and runoff as opposed to changes in nonpoint sources. Estuary water quality models did not identify significant long-term changes within Lavaca Bay for dissolved oxygen or chlorophyll-*a*. However, site specific long-term increases in both organic and inorganic nitrogen are concerning. Further analysis found freshwater inflow was a strong driver of nutrient and chlorophyll-*a* concentrations within Lavaca Bay but there was no evidence that changes in nutrient loading explained variation in dissolved oxygen or chlorophyll-*a* concentrations. In addition to providing a baseline assessment of watershed nutrient loading and water quality responses in the Lavaca Bay watershed, this study provides methodological support for the use of semiparametric methods in load regression models and estuary assessments.

bibliography: bibliography.bib
output: rticles::springer_article
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      out.extra = "")  # This forces knitr to label all figures.

library(targets)
library(knitr)
library(tidyverse)
library(kableExtra)
library(patchwork)
library(ggrepel)
library(rcartocolor)
library(twriTemplates)
library(ggspatial)
library(ragg)
library(modelsummary)
library(scico)
library(ggridges)
library(ggtext)
library(lubridate)
library(units)
library(gratia)
library(gt)

## this sets our default code chunk options
knitr::opts_chunk$set(echo = FALSE,
                      ## figure chunk options
                      dev = "ragg_png",
                      dpi = 300,
                      fig.align = "center",
                      out.width = "100%")

@@

# Introduction {#sec1}

@@

# Methods {#sec2}

## Location and Data
Lavaca Bay is 190 km^2^ with the majority of freshwater inflow provided by the 
Lavaca and Navidad River systems (Figure\ \ref{fig:fig1}). The Garcitas-Arenosa, 
Placedo Creek, and Cox Bay watersheds provide additional freshwater inflows. 
The entire watershed land area is 8,149 km^2^ and primarily rural. Watershed 
land cover and land use is 50% grazed pasture and rangeland, 20% cultivated 
cropland (primarily rows crops such as corn, cotton, and sorghum), and 5% 
suburban/urban. Pasture and rangeland is concentrated in the Lavaca River 
watershed, while cultivated crops are generally located along the eastern 
tributaries of the Navidad river. The Lavaca and Navidad River watersheds are a 
combined 5,966 km^2^, or approximately 73% of the entire Lavaca Bay watershed 
area. Discharge from the Navidad River is regulated by Lake Texana which has 
been in operation since 1980. Lake Texana provides 0.210 km^3^ of water 
storage and discharges into the tidal section of the Navidad River which 
ultimately joins the tidal section of the Lavaca River 15 km upstream of the 
confluence with the Lavaca Bay.

```{r fig1, fig.width=6.75, fig.width=5.2, out.width="5.2in", fig.cap="Map of Lavaca Bay and the contribution watershed. The freshwater sites are the most downstream freshwater stream locations with water quality and streamflow data used for nutrient load models. Water quality concentration data at the three Lavaca Bay sites were used to assess relationships between freshwater flows, loads and estuary water quality."}

knitr::include_graphics("../fig1.png")

@@

Daily discharges for the Lavaca River (USGS-08164000, Figure\ \ref{fig:fig1}) 
were obtained from the United States Geologic Survey (USGS) National Water 
Information System using the *dataRetrieval* R package 
[@deciccoDataRetrievalPackagesDiscovering2022]. Gaged daily discharges from 
the outlet of Lake Texana on the Navidad River (USGS-0816425) were provided by 
the Texas Water Development Board (TWDB) 
(April 21, 2022 email from R. Neupane, TWDB).

Water quality sample data for the two freshwater and three estuary locations were 
obtained from the Texas Commission on Environmental Quality (TCEQ) Surface Water 
Quality Monitoring Information System. Data submitted through the system 
are required to be collected under Quality Assurance Project Plans and lab 
method procedures outlined by the TCEQ's procedures manual. 
The QAPP and procedures manuals ensure the consistent collection and laboratory 
methods are applied between samples collected by different entities and under 
different projects. 
All sites had varying lengths of and availability of data.
For freshwater locations, TP from January 2000 through December 2020 and 
nitrate-nitrogen (NO~3~) data from January 2005 through December 2020 were 
downloaded (Table\ \ref{tab:table1}). Less than 5-years of total
nitrogen and TKN concentration data were available at the freshwater sites and
deemed insufficient to develop load estimation models 
[@horowitzEvaluationSedimentRating2003; @snelderEstimationCatchmentNutrient2017]. 
The three estuary sites included an upper Lavaca Bay site near the outlet of the
Lavaca River system (TCEQ-13563), a mid-Lavaca Bay site (TCEQ-13383), and the 
lower Lavaca Bay site near the mouth of the Bay (TCEQ-13384). 
For estuary locations, we obtained data for TP, Nitrite+Nitrate (NO*~x~*), TKN, 
chlorophyll-*a*, and DO concentrations from January 2005 through December 2020 
(Table\ \ref{tab:table2}).

```{r table1}

@@

datasummary(`Station ID` * (`TP (mg/L)` + `NO3 (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
               Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
            data = df, output = "kableExtra", format = "latex",
            caption = "Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.")

```{r table2}
#| warning: false

@@

modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                           `NOx (mg/L)` +
                           `TKN (mg/L)` +
                           `Chlorophyll-a (ug/L)` +
                           `DO (mg/L)`) ~
                           Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                         data = df, output = "kableExtra", format = "latex",
                         caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.")
```

@@

Estimates of NO~3~ and TP loads at the Lavaca River (USGS-08164000) and the 
outlet of Lake Texana on the Navidad River (USGS-08164525) were developed using 
GAMs relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to predict
nutrient concentrations for each day in the study period. GAMs were fit using 
the *mgcv* package in R which makes available multiple 
types of smooth functions with automatic smoothness selection [@woodFastStableRestricted2011]. 
The general form of the model related NO~3~ or TP concentration to a long term
tend, season, streamflow, and two different antecedent discharge terms, 
shown in Eq. \ref{eq1}:

\begin{align}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(log1p(Q)) + f_4(ma) + f_5(fa), \nonumber \\
y &\sim \mathcal{N}(\mu,\,\sigma^{2})
\label{eq1}
\end{align}

where $\mu$ is the conditional expected NO~3~ or TP concentration, $g()$ is the 
log-link, $\alpha$ is the intercept, $f_n()$ are smoothing functions. $y$ is the 
response variable (NO~3~ or TP concentration) modeled as normally distributed 
with mean $\mu$ and standard deviation $\sigma$. *ddate* is the date converted to 
decimal notation, *yday* is numeric day of year (1-366), and *log1p(Q)* is the 
natural log of mean daily streamflow plus 1.

Moving average (*ma*) is an exponentially smoothed moving average that attempts 
to incorporate the influence of prior streamflow events on concentration at the 
current time period [@wangLoadEstimationUncertainties2011;  
@kuhnertQuantifyingTotalSuspended2012; @zhang_improving_2017], using Eq. \ref{eq2}:

\begin{align}
ma(\delta) &= d{\kappa_{i-1}}+(1-\delta)\hat{q}_{i-1}, \nonumber \\
\kappa_{i} &=\sum_{m=1}^{i}\hat{Q}_m
\label{eq2}
\end{align}

where $\delta$ is the discount factor (here, set equal to 0.95), $\kappa_i$
is the cumulative flow (*Q*) up to the *i*th day.

Flow anomaly (*fa*) is a unitless term that represents how wet or dry the 
current time period is from a previous time period [@vecchiaTrendsConcentrationsUse2009; 
@zhang_improving_2017]. Long-term flow anomaly (*ltfa*) is the streamflow over 
the previous year relative to the entire period [@zhang_improving_2017] and 
calculated using Eq. \ref{eq3}:

\begin{equation}
ltfa(t) = \bar{x}_{1\,year}(t) - \bar{x}_{entire\,period}
\label{eq3}
\end{equation}

and the short-term flow anomaly (*stfa*) calculated as the current day flow 
compared to the preceding 1-month streamflow using \ref{eq4}:

\begin{equation}
stfa(t) = x_{current\,day}(t) - \bar{x}_{1\,month}(t)
\label{eq4}
\end{equation}

where $x$ are the averages of log-transformed streamflow over the antecedent 
period (*1-year*, *1-month*, etc.) for time *t*.  We used *ltfa* in NO~3~ models 
and *stfa* in TP models based on previous work demonstrating major improvements 
in NO~x~ regression models that incorporated 
*ltfa* and moderate improvements in TP regression models that incorporated 
*stfa* [@zhang_improving_2017].

@@

To test if changes in freshwater inflow and nutrient loading had explanatory
effect on changes in estuary water quality a series of GAM models were fit at
each site relating parameter concentration to temporal trends ([@eq-eq5]), 
temporal trends and inflow (@eq-eq6), and temporal trends, inflow, and
nutrient loads (@eq-eq7):
\begin{align}
g(\mu) &= \alpha + f_1(ddate) + f_2(yday), \label{eq5} \\
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(Q), \label{eq6} \\
g(\mu) &= \alpha + f_1(ddate) + f_2(yday) + f_3(Q) + f_4(Load) \label{eq7}
\end{align}

where $\mu$ is the conditional expected response (nutrient concentration), $g()$ 
is the log link, and response variable was modeled as Gamma distributed with 
mean $\mu$ and scale $\lambda$. $f_1(ddate)$ is decimal date smoothed with a 
thin-plate regression spline, $f_2(yday)$ is the numeric day of year smoothed 
with a cyclic cubic regression spline, $f_3(Q)$ is mean daily inflow (the 
combined measurements from Lavaca River and Navidad River) and $f_4(Load)$ is 
the total NO~3~ or TP watershed load. The set of models specified for each water
quality response are in Table\ \ref{tab:table3}.

```{r table3}
#| warning: false
df <- tibble(
  `Parameter` = c(rep("TP", 3), rep("NO\\textsubscript{x}", 3),
               rep("Chlorophyll-\\textit{a}", 3), rep("Dissolved Oxygen", 3),
               rep("TKN", 2)),
  Model = c("Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow", "Flow+Load",
            "Temporal", "Flow"),
  `Model Terms` = c("s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(NO\\textsubscript{3} Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO\\textsubscript{3} Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)",
                    "s(ddate) + s(yday) + s(Q) + s(TP Load) + s(NO\\textsubscript{3}  Load)",
                    "s(ddate) + s(yday)",
                    "s(ddate) + s(yday) + s(Q)")
)

kbl(df, format = "latex",
    caption = "Set of GAM models specified for each water quality parameter response.",
    booktabs = TRUE,
    escape = FALSE) |> 
  collapse_rows(columns = 1, latex_hline = "major")

# flextable::flextable(df) |> 
#   set_caption(caption = "Set of GAM models specified for each water quality parameter response.",
#               autonum = run_autonum(seq_id = "tab", bkm = "table3")) |> 
#   style(pr_t = fp_text_default(font.size = 9), part = "body") |>
#   style(pr_t = fp_text_default(font.size = 9, bold = TRUE), part = "header") |>
#   set_table_properties(layout = "autofit") |>
#   padding(padding.top = 0, padding.bottom = 0, part = "body")

@@

Because streamflow and nutrient loads are tightly correlated, freshwater inflow
can mask signals from nutrient loads alone. Following the 
methodology implemented by @murphyNutrientImprovementsChesapeake2022,
both streamflow and nutrient loads were prepossessed to account for season
and flow. Freshwater inflow and nutrient loads were replaced by 
seasonally adjusted log transformed inflow and flow-adjusted 
log transformed nutrient loads obtained by fitting a GAM relating season 
(day of year) to log transformed daily freshwater inflow values
and a GAM relating log transformed NO~3~ or TP loads to log transformed daily
inflow.Response residuals from the respective GAM models were 
used as *Q* and *Load* in Eq.\ \ref{eq6} and Eq.\ \ref{eq7}.

@@

# Results {#sec3}

@@

Using evaluation criteria recommended by @moriasiHydrologicWaterQuality2015, 
predictive performance of nutrient loading GAMs ranged from "satisfactory" to
"very good" based on median NSE, *r*, and PBIAS metrics calculated using 5-fold 
cross validation. Median goodness-of-fit metrics for NO~3~ models in the Lavaca 
River were 0.34 NSE, 0.70 *r*, and 2.00 PBIAS. Navidad River NO~3~ models 
appeared to perform slightly better with 0.48 NSE and 0.87 *r* but with higher 
bias at 10.90 PBIAS. Generally, TP models performed better than NO~3~ models. 
Median goodness-of-fit metrics for TP in the Lavaca River were 0.81 NSE, 
0.93 *r*, and -7.20 PBIAS. Navidad River TP models has similar performance 
with 0.91 NSE, 0.99 *r*, and -3.30 PBIAS. Density plots of
metrics show similar distribution of values between sites for the 
same parameter, with the exception *r* values for NO~3~ loads where 
Lavaca River had a much larger variance in values compared to the Navidad River
(Figure\ \ref{fig:fig2}). TP GAMS had higher average NSE and *r* values and 
lower variance in metric values compared to NO~3~.

```{r fig2, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.65, fig.cap="Density plots of goodness-of-fit metrics (NSE, \\textit{r}, and PBIAS) from repeated 5-fold cross validation between predicted nutrient loads from GAM models and measured nutrient loads. Color indicates the tail probability calculated from the empirical cumulative distribution of the goodness-of-fit metrics."}

@@

Predicted annual NO~3~ and TP loads show considerable variation, generally following
patterns in discharge (Figures\ \ref{fig:fig3}, \ref{fig:fig4}). 
Flow-normalized TP loads at both sites and flow-normalized NO~3~ loads in the Lavaca River
indicated watershed based loads did not change much over time
when accounting for variation driven by streamflow (Figure\ \ref{fig:fig3}). 
Flow-normalized loads in the Lavaca River showed small variation over time with 
some decreases in NO~3~ loads since 2013.

```{r fig3, error=FALSE, warning=FALSE, message=FALSE, fig.width=6.85, fig.asp=0.7, fig.cap="Aggregated estimated annual and flow-normalized annual NO~3~ and TP loads for USGS-08164000 and USGS-08164525."}

load <- tar_read(daily_no3_08164000, store = store)
fn_load <- tar_read(daily_no3_08164000_fn, store = store)
